---
title: 'BMI 715 Final Project'
author: "Dan Kats"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Set-up
```{r, message=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)

nhanes <- read.csv("nhanes_13_14_subset.csv")
nhanes$X <- NULL
```

## Exploratory analysis and QC

##### Relationship between variables
```{r}

```

##### Correlations/collinearity between predictors
```{r}

```

##### Scale predictors that have a different range
```{r}

```

##### Code categorical variables appropriately
```{r}

```

##### Transform data if needed
```{r}

```

##### Assess/remove/impute missing values
```{r}

```


## Regression/classification model: Running

##### Evaluate assumptions for your chosen model
```{r}

```

##### Variable selection (forward, backward, stepwise), regularization (LASSO, Ridge, ElasticNet), and/or train/test splitting
```{r}

```

##### Run your chosen model and report any key statistics
```{r}

```


## Regression/classification model: Evaluating

##### Choose, calculate, and appropriately interpret a metric
```{r}

```

##### Include 1+ plot that you use to evaluate model fit
```{r}

```


## Compare your model to one with different variables

##### Run the alternate model
```{r}

```

##### Calculate a metric to compare the models 
```{r}

```